Prosper Loans Exploration by Paron Sarampakhul
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(scales)
library(psych)
library(car)
library(RColorBrewer)
library(memisc)
```

```{r echo=FALSE,message=FALSE, warning=FALSE,Load_the_Data}
# Load the Data
loan_data <- read.csv(file="prosperLoanData.csv", header=TRUE, sep=",")
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Overview_dimension}
dim(loan_data)

```

```{r echo=FALSE,message=FALSE, warning=FALSE,Overview_structure}
str(loan_data)
```

```{r echo=FALSE,message=FALSE, warning=FALSE,Overview_summary}
summary(loan_data)
```

> The dataset contains 81 variables with 113,937 observations.


# Univariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE,Univariate_Plots1}
ggplot(data=loan_data, aes(as.Date(ListingCreationDate))) +
  geom_bar() +
  scale_x_date(date_breaks='6 months', labels = date_format("%Y-%b")) +
  labs(title="Histogram for Listing Creation Date") +
  labs(x="ListingCreationDate", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

> I see that from the dataset that there are 2 columns for credit grades,\
pre-2009 and after July 2009, therefore I want to see the number of listings\
over time. It turns out that there is a gap between late 2008 and mid 2009 \
where there are no listings. I wonder whether the gap has something to do \
with the Financial Crisis in late 2008. There is an increasing trend for \
number of listings from 2009 to 2013. This could be because the economy \
started to recover.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots2}
ggplot(data=subset(loan_data,as.character(loan_data$CreditGrade)!=""),
       aes(CreditGrade)) +
  geom_bar() +
  labs(title="Histogram for Credit Grade (Pre-2009)") +
  labs(x="Credit Grade", y="Count") +
  scale_x_discrete(limits=c("AA","A","B","C","D","E","HR","NC")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots3}
ggplot(data=subset(loan_data,as.character(loan_data$ProsperRating..Alpha.)!=""),
       aes(ProsperRating..Alpha.)) +
  geom_bar() +
  labs(title="Histogram for ProsperRating (Alpha) (Post-2009)") +
  labs(x="Propsper Rating (Alpha)", y="Count") +
  scale_x_discrete(limits=c("AA","A","B","C","D","E","HR","NC")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> I want to see if the distributions of credit ratings are the same for\
pre-2009 listings and post-2009 listings. I have to re-arrange the ratings from\
best to worse for the x-axis. Pre-2009 listings have high numbers of B-rated,\
C-rated, and D-rated ratings. The number of AA-rated listings is also more than\
the number of A-rated listings. Post-2009 listings have high numbers of\
A-rated, B-rated, C-rated, and D-rated listings. The number of AA-rated\
listings is significantly lower than the number of A-rated listings.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots4}
ggplot(data=loan_data, aes(loan_data$LoanStatus)) +
  geom_bar() +
  labs(title="Histogram for Loan Status") +
  scale_x_discrete(limits=c("Cancelled","Completed","Current",
                            "FinalPaymentInProgress","Past Due (1-15 days)",
                            "Past Due (16-30 days)","Past Due (31-60 days)",
                            "Past Due (61-90 days)","Past Due (91-120 days)", 
                            "Past Due (>120 days)","Defaulted","Chargedoff")) +
  labs(x="Loan Status", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> I want to see the status of the loans. I have to re-arrange the status of\
the loans that are past due.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots5}
ggplot(data=subset(loan_data,as.character(loan_data$ProsperScore)!=""),
       aes(ProsperScore)) +
  geom_histogram() +
  labs(title="Histogram for ProsperRating (Alpha)") +
  scale_x_continuous(breaks = seq(1,11,by=1)) +
  labs(x="Credit Grade", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


> I want to see the distribution of ProsperScore, which is a custom risk score \
built using historical Prosper data. From the provided data variables\
definitions, the score should be from 1 to 10, with 10 being the best score.\
I wonder why there are some listings with score equal to 11.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots6}
ggplot(data=loan_data, aes(factor(Term))) +
  geom_bar() +
  labs(title="Histogram for Term") +
  labs(x="Term (Months)", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots7}
table(loan_data$Term)
```

> There are only 3 available terms for loans: 12 months, 36 months and 60\
months. Most listings are due in 36 months.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots8}
ggplot() +
  geom_histogram(data=loan_data, aes(BorrowerAPR,color="BorrowerAPR"),
                 fill = "red", alpha = 0.2) +
  labs(title="Histogram BorrowerAPR") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  scale_colour_manual("Rate", values = c("red", "blue")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots8_2}
ggplot() +
  geom_histogram(data=loan_data, aes(BorrowerRate,color="BorrowerRate"),
                 fill = "blue", alpha = 0.2) +
  labs(title="Histogram & BorrowerRate") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  scale_colour_manual("Rate", values = c("red", "blue")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots9}
summary(loan_data$BorrowerAPR)
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots10}
summary(loan_data$BorrowerRate)
```

>Both BorrowerAPR and Borrowerate have normal distribution. Mean Borrower APR
is 21.88% and mean borrower rate is 19.28%. This makes sense because APR is a
broader measure of the cost of a mortgage. It includes the interest rate
(borrower rate) plus other costs such as broker fees, discount points and some
closing costs
(https://www.bankrate.com/finance/mortgages/apr-and-interest-rate.aspx).


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots11}
ggplot(data=loan_data, aes(LenderYield)) +
  geom_histogram() +
  labs(title="Histogram for Lender Yield") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots12}
ggplot(data=loan_data, aes(EstimatedEffectiveYield)) +
  geom_histogram() +
  labs(title="Histogram for Estimated Effective Yield") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots13}
ggplot(data=loan_data, aes(EstimatedLoss)) +
  geom_histogram() +
  labs(title="Histogram for Estimated Loss") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```
```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots14}
ggplot(data=loan_data, aes(EstimatedReturn)) +
  geom_histogram() +
  labs(title="Histogram for Estimated Return") +
  labs(x="Rate", y="Count") +
  scale_x_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots15}
summary(loan_data$LenderYield)
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots16}
summary(loan_data$EstimatedEffectiveYield)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots17}
summary(loan_data$EstimatedLoss)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots18}
summary(loan_data$EstimatedReturn)
```

> Mean lender yield (interest rate less servicing fee) = 18.27%, mean \
estimated effective yield = 16.9%, mean estimated loss = 8%, mean estimated \
return = 9.6%. Lender yield, estimated effective yield and estimated loss \
have normal distribution but with a small peak on the right side. It seems \
that the higher returns are associated with higher risk of loss. The net \
estimated return therefore has normal distribution without a small peak \
on the right side.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots19}
ggplot(data=subset(loan_data,as.character(loan_data$ListingCategory..numeric.)!=""),
       aes(ListingCategory..numeric.)) +
  geom_bar() +
  labs(title="Histogram for Listing Category") +
  scale_x_continuous(breaks=seq(0,20,by=1)) +
  labs(x="Category", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> Majority of the loan category is debt consolidation.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots20}
ggplot(data=subset(loan_data,as.character(loan_data$Occupation)!=""),
       aes(Occupation)) +
  geom_bar() +
  labs(title="Histogram for Occupation") +
  labs(x="Occupation", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


> The histogram for occupation shows 2 significant peaks: Other and \
Professional. This is because the term 'Professional' is a broard term,\
and 'Other' is a blanket term for other occupations that may not be listed \
in the survey.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots21}
ggplot(data=subset(loan_data,as.character(loan_data$EmploymentStatus)!=""),
       aes(EmploymentStatus)) +
  geom_bar() +
  labs(title="Histogram for Employment Status") +
  labs(x="Employment Status", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> The histogram for employment status shows that most borrowers are employed.\
I wonder what are the differnece between 'Employed' and 'Full-time'.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots22}
ggplot(data=subset(loan_data,as.character(loan_data$EmploymentStatusDuration)!=""),
       aes(EmploymentStatusDuration)) +
  geom_histogram() +
  scale_x_log10(breaks= c(2,5,10,20,50,100,200,500,1000)) +
  xlab('Months') +
  labs(title="Histogram for Employment Status Duration") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots23}
summary(loan_data$EmploymentStatusDuration)
```


> Histogram for employment duration shows that majority of the borrowers have\
been employed less than 6 years. The median duration is 67 months.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots24}
ggplot(data=subset(loan_data,as.character(loan_data$IsBorrowerHomeowner)!=""),
       aes(IsBorrowerHomeowner)) +
  geom_bar() +
  labs(title="Histogram for Homeowner Status") +
  labs(x="Homeowner Status", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> The portion of borrowers who are homeowners and the portion of borrowers who are not homeowners are close.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots25}
ggplot(data=subset(loan_data,as.character(loan_data$IncomeRange)!=""), 
       aes(IncomeRange)) +
  geom_bar() +
  labs(title="Histogram for Income Range") +
  scale_x_discrete(limits=c("Not displayed","Not employed","$0","$1-24,999",
                            "$25,000-49,999","$50,000-74,999","$75,000-99,999",
                            "$100,000+")) +
  labs(x="Income Range", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots26}
summary(loan_data$StatedMonthlyIncome * 12)
```


> The income range of borrowers are left-skewd normally distributed, with\
mean stated annual income = $67,296 and median stated annual income = $56,000.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots27}
ggplot(data=subset(loan_data,loan_data$CreditScoreRangeLower!=""), 
       aes((CreditScoreRangeLower + CreditScoreRangeUpper)/2)) +
  geom_bar() +
  labs(title="Histogram for Credit Score") +
  labs(x="Credit Score", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


> Since the lower range and upper range of credit score provided by a \
consumer credit rating agency, I take the mid points and plot them. The \
credit score of the borrowers are left-skewed normally distributed. I suspect \
that one of the factors used to determine a credit score is the income.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots28}
ggplot(data=subset(loan_data,loan_data$CurrentCreditLines!=""), 
       aes(CurrentCreditLines)) +
  geom_bar() +
  labs(title="Histogram for Current Credit Lines") +
  labs(x="Number of Credit Lines", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots29}
ggplot(data=subset(loan_data,loan_data$OpenCreditLines!=""), 
       aes(OpenCreditLines)) +
  geom_bar() +
  labs(title="Histogram for Open Credit Lines") +
  labs(x="Number of Open Credit Lines", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots30}
ggplot(data=subset(loan_data,loan_data$OpenRevolvingAccounts!=""), 
       aes(OpenRevolvingAccounts)) +
  geom_bar() +
  labs(title="Histogram for Open Revolving Accounts") +
  labs(x="Number of Accounts", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots31}
summary(loan_data$CurrentCreditLines)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots32}
summary(loan_data$OpenCreditLines)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots33}
summary(loan_data$OpenRevolvingAccounts)
```

> Number of current credit lines, number of open credit lines, and number of\
open revolving accounts are right-skewed normally distributed. I wonder whether\
these are correlated with the borrower's credit score, the listing's credit\
grade, and listing's estimated loss.


```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots34}
ggplot(data=subset(loan_data,loan_data$OpenRevolvingMonthlyPayment!=""),
       aes(OpenRevolvingMonthlyPayment)) +
  geom_histogram() +
  labs(title="Histogram for Open Revolving Monthly Payments") +
  scale_x_log10(breaks= c(10,20,50,100,200,500,1000,2000,5000,10000))+
  labs(x="USD", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots35}
ggplot(data=subset(loan_data,loan_data$CurrentDelinquencies!=""), 
       aes(CurrentDelinquencies)) +
  geom_histogram() +
  scale_x_continuous() +
  xlim(0,20) +
  ylim(0,15000) +
  labs(title="Histogram for Current Delinquencies") +
  labs(x="Number of Delinquencies", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots36}
ggplot(data=subset(loan_data,loan_data$AmountDelinquent!=""),
       aes(AmountDelinquent)) +
  geom_histogram() +
  scale_x_log10(breaks= c(10,20,50,100,200,500,1000,2000,5000,10000,20000,
                          50000,100000)) +
  labs(title="Histogram for Amount Delinquent") +
  labs(x="USD", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots37}
ggplot(data=subset(loan_data,loan_data$LoanOriginalAmount!=""),
       aes(LoanOriginalAmount)) +
  geom_histogram() +
  scale_x_log10(breaks= c(10,20,50,100,200,500,1000,2000,5000,10000,20000,
                          50000,100000)) +
  labs(title="Histogram for Original Loan Amount") +
  labs(x="USD", y="Count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Univariate Analysis

### What is the structure of your dataset?
The data set consists of 81 variables and 113,937 records. Each record contains
the information of the loan listing, which include loan key, interest rate,
estimated return rate, estimated loss rate, loan amount, loan category, credit
rating, borrower information, etc.

### What is/are the main feature(s) of interest in your dataset?
The main features in the dataset are borrower rate, estimated effective yield,
estimated loss rate, estimated return rate, and credit rating of a loan. I
would like to find out what features determine the mentioned values. I suspect
that the main factors should include the annual income of borrower, employment
status, credit score, number of open credit lines, number of open revolving
accounts, current delinquencies, and amount delinquent.

I suspect that the high annual income and employed status will result in higher
credit score. The small number of credit lines, revolving accounts, and
delinquencies will also result in higher credit score. Higher credit score
should result in lower borrower rate, lower estimated loss, and better credit
rating of the listing.

### What other features in the dataset do you think will help support your nvestigation into your feature(s) of interest?
Other factors that may determine the credit score include the size of the loan
orignal amount, employment duration, TotalProsperPaymentsBilled,
%OnTimeProsperPayments vs %LateProsperPayments, ProsperPrincipalOutstanding,
and number of recommendations the borrower receives.

### Did you create any new variables from existing variables in the dataset?
I only transformed the monthly stated income to annual by multiplying with 12
as I would like to see whether the monthly stated income tallies with the
column IncomeRange in terms of their distributions.

I also find the mid point of borrower's credit score by summing lower range and
upper range, and then dividing by 2.

### Of the features you investigated, were there any unusual distributions?

I log-transformed the right skewed of employment durations, revolving monthly
payments, amount delinquent, and the size of orignal loan amount.


# Bivariate Plots Section


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots1}
cor(loan_data[c("BorrowerRate","EstimatedLoss","EstimatedReturn",
                "ProsperRating..numeric.","ProsperScore",
                "EmploymentStatusDuration","CreditScoreRangeLower",
                "CreditScoreRangeUpper","OpenCreditLines",
                "OpenRevolvingAccounts","OpenRevolvingMonthlyPayment",
                "CurrentDelinquencies","AmountDelinquent","StatedMonthlyIncome",
                "TotalProsperPaymentsBilled","OnTimeProsperPayments",
                "LoanOriginalAmount","Recommendations")], use = "complete.obs")

```

> There is a high correlation between BorrowerRate and Estimated Loss.\
This makes sense since investors should require higher rates for riskier assets.\
As expected, the BorrowerRate is negatively correlated with ProsperScore,\
ProsperRating and CreditRating. I wonder what factors are used to determine \
the interest rate for each listing.

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots2}
ggplot(aes(x=StatedMonthlyIncome,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.02) +
  xlim(0,20000) +
  xlab('$') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> From the scatterplot, it is not very clear whether StatedMonthlyIncome is \
correlated with the BorroweRate. Let's try a boxplot

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots3}

ggplot(data=subset(loan_data,as.character(IncomeRange)!=""),
       aes(x=IncomeRange,y=BorrowerRate)) +
  geom_boxplot() +
  scale_x_discrete(limits=c("Not displayed","Not employed","$0","$1-24,999",
                            "$25,000-49,999","$50,000-74,999","$75,000-99,999",
                            "$100,000+")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots4}

describeBy(loan_data$BorrowerRate, group=loan_data$IncomeRange, 
           mat=TRUE)[c(7,8,1,2,4,5,6,3),]

```


> From the boxplot, it is as expected that the higher the income the borrower\
has, the less risky the loan is. However, it is strange that the median\
BorrowerRate for the group whose IncomeRange = $0 is lower than those with \
IncomeRange $1-74,999.


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots5}

ggplot(data=subset(loan_data,as.character(EmploymentStatus)!=""),
       aes(x=EmploymentStatus,y=BorrowerRate)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots6}

describeBy(loan_data$BorrowerRate, group=loan_data$EmploymentStatus, mat=TRUE)

```

> As expected, the median BorrowerRate for borrowers who are employed are\
lower than those who are not employed. However, I am curious why the\
median BorrowerRate for borrowers who are part-timers are lower than \
those who are full-timers.


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots7}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=loan_data) +
  geom_jitter(alpha=0.01)
```

> From the scatterplot, it is clear that the higher the credit score the \
borrower has, the lower the BorrowerRate is required.

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots8}

ggplot(data=subset(loan_data,as.character(ProsperRating..Alpha.)!=""),
       aes(x=ProsperRating..Alpha.,y=BorrowerRate)) +
  geom_boxplot() +
  scale_x_discrete(limits=c("AA","A","B","C","D","E","HR"))

```


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots9}

describeBy(loan_data$BorrowerRate, group=loan_data$ProsperRating..Alpha.,
           mat=TRUE)[c(1,3,2,4,5,6,7,8),]

```

> This is as expected since I suspect that the credit score of the borrower\
will determine the credit rating of the loan, and the worse loan rating will \
require the higher interest rate.


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots10}
ggplot(data=subset(loan_data,Term=!""),aes(x=as.factor(Term),y=BorrowerRate)) +
  geom_boxplot()+
  xlab('Term (Months)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots11}

describeBy(loan_data$BorrowerRate, group=loan_data$Term, mat=TRUE)

```

> As expected, the shorter term (12-month) loan has lower median BorrowerRate than the longer term (36 and 60 month) has, since the longer-term loans are more exposed to interest rate risks. However, I am surprised that the difference between the BorrowerRate of 36-month loan and 60-month loan is small.


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots12}
ggplot(data=subset(loan_data,ListingCategory..numeric.!=""),
       aes(x=as.factor(ListingCategory..numeric.),y=BorrowerRate)) +
  geom_boxplot() +
  xlab('Listing Category')
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots13}

describeBy(loan_data$BorrowerRate, 
           group=loan_data$ListingCategory..numeric., mat=TRUE)

```

> Differnt categories of loans can have different BorrowerRate,\
but I will not go deeper.

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots14}
ggplot(aes(x=OpenCreditLines,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots15}
ggplot(aes(x=OpenRevolvingAccounts,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots16}
ggplot(aes(x=CurrentDelinquencies,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01) +
  xlim(0,20)
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots17}
ggplot(aes(x=AmountDelinquent,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01) +
  xlim(0,10000) +
  xlab('Amount Delinquent ($)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots18}
ggplot(aes(x=
PublicRecordsLast10Years,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01) +
  xlim(0,10)
```


> There is no clear correlation between number of credit lines/open revolving\
accounts and BorrowerRate. However, the number of delinquencies/public records\
is positively correlated with the BorrowerRate.


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots19}
ggplot(aes(x=
LoanOriginalAmount,y=BorrowerRate), data=loan_data) +
  geom_jitter(alpha=0.01) +
  xlab('Loan Original Amount ($)')
```

> The size of the loan does not seem to be correlated with the BorrowerRate.


```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots20}
summary(lm(formula = BorrowerRate ~ StatedMonthlyIncome, data=loan_data))
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots21}
summary(lm(formula = BorrowerRate ~ CreditScoreRangeLower, data=loan_data))
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots22}
summary(lm(formula = BorrowerRate ~ EmploymentStatus, data=loan_data))
```

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots23}
summary(lm(formula = BorrowerRate ~ CurrentDelinquencies, data=loan_data))
```

> The R-Squared values show that it is not sufficient to explain the borrower \
rate with just one variable.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The BorrowerRate is positively and strongly correlated with the Prosper Credit
Rating. It is likely that the rate ranges are determined by the loan credit
rating.

Other features that are positively, though less strongly correlated with
BorrowerRate, include staetd monthly income, borrower's credit score, number of
delinquents, and number of public records in the past 10 years.

The features that do not seem to be correlated with BorrowerRate are number of
open credit lines, number of open revolving accounts, and size of the loan.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)? The interest rate also depends on the
term of the loan. Shorter term loans have lower required interest rates than
longer term loans do. Another interesting observation is that the median
interest rate of the loan category 10 (Cosmetic Operation) is significantly
higher than that of loan category 4 (Personal Loan), 25% vs 15%.

### What was the strongest relationship you found?
The BorrowerRate and the Prosper Credit Rating of the listing are positively
and strongly correlated. This is expected as the rating is classified by its
riskiness. The more interesting question is how do we determine the riskiness
of each listing.


# Multivariate Plots Section


```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots1}
ggplot(aes(x=
(ProsperPaymentsLessThanOneMonthLate+
ProsperPaymentsOneMonthPlusLate)/TotalProsperPaymentsBilled,y=BorrowerRate),
data=loan_data) +
  geom_jitter(alpha=0.01) +
  xlim(0,0.3) +
  xlab('%Late Prosper Payments')
```


> The Prosper settlement records do not seem to be correlated to the BorrowerRate.

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots2}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=as.factor(Term))) +
  scale_color_brewer(type = 'qual') +
  xlab('Credit Score')
```

> Short term listings (12-month) clearly have lower BorrowRate. Also, as the\
credit score increases, the BorrowRate for short-term listings seem to decrease.\
However, for the longer term listings (36-month and 60-month), the variance is\
high. It is not clear which one has lower BorrowRate, and whether credit\
score matters.


```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots3}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=as.factor(EmploymentStatus)), alpha = 0.5) +
  scale_color_brewer(type = 'qual') +
  xlab('Credit Score')
```

> This plot is crowded by Employed status. It does not tell anything much.

```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots4}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=ProsperRating..Alpha.), alpha = 0.1) +
  scale_color_brewer(type = 'seq',palette='Reds',limits=c("AA","A","B","C",
                                                          "D","E","HR")) +
  xlab('Credit Score')
```

> BorrowerRate are clearly layered by ProsperRating..Alpha. The BorrowerRate \
also has moderate negative correlation with the credit score of the borrower.


```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots5}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=ProsperRating..Alpha.)) +
  scale_color_brewer(type = 'seq',palette='Reds',limits=c("AA","A","B","C",
                                                          "D","E","HR")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Term) + 
  xlab('Credit Score')

```

> Shorter term listings clearly have lower BorrowerRate. The rates are further \
layered by ProsperRating..Alpha. We can also see moderate negative correlation \
between BorrowerRate and CreditScore.


```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots6}
ggplot(aes(x=StatedMonthlyIncome,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=ProsperRating..Alpha.)) +
  scale_color_brewer(type = 'seq',palette='Blues',limits=c("AA","A","B","C","D",
                                                           "E","HR")) +
  xlim(0,30000) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Term) +
  xlab('Stated Monthly Income ($)')

```


```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots7}
m1 <- lm(formula = I(BorrowerRate) ~ CreditScoreRangeLower, 
         data = subset(loan_data,,as.character(ProsperScore)!=""))
m2 <- update(m1, ~. + StatedMonthlyIncome)
m3 <- update(m2, ~. + CurrentDelinquencies)
m4 <- update(m3, ~. + ProsperRating..Alpha.)
m5 <- update(m4, ~. + Term)
m6 <- update(m5, ~. + EmploymentStatus)
mtable(m1,m2,m3,m4,m5,m6)
```


The variables in this linear model can account for 76.4% of the variance in the
BorrowerRate of the loan listings. The most important factor is the credit
rating of the loan, followed by credit score of the borrower. Other variables
poorly contribute to the prediction.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

ProsperRating..Alpha clearly determine the ranges of the BorrowRate. The
BorrowerRate also has moderate negative correlation with the credit score of
the borrower. The stated monthly income has weaker correlation to the borrower
rate.

The 12-month clearly has lower borrower rate than those of 36-month and
60-month loans.

### Were there any interesting or surprising interactions between features?

Yes, I am surprised that the monthly stated income does not matter much in
determining the borrower rates. Also, I would have thought that the 60-month
loans would have higher interest rates than those of the 36-month loans since
the former are more exposed to the interest rate change risk.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created models to predict a loan's borrower rate from borrower's credit
score, stated monthly income, Prosper credit rating of the listing, loan term,
and borrower's employment status. These variables can account for 76.4% of the
variance in the BorrowerRate. However, I still cannot find determining factors
for credit rating of the loan listings.

------

# Final Plots and Summary


### Plot One

```{r echo=FALSE,message=FALSE, warning=FALSE, Summary1}

ggplot(data=subset(loan_data,as.character(ProsperRating..Alpha.)!=""),
       aes(x=ProsperRating..Alpha.,y=BorrowerRate)) +
  geom_boxplot() +
  scale_x_discrete(limits=c("AA","A","B","C","D","E","HR"))

```

### Description One
> The loans are categorized by its riskiness from low risk to high risk (AA -> HR). The riskier the loans are, the higher interest rates they require.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary2}
ggplot(aes(x=(CreditScoreRangeLower+CreditScoreRangeUpper)/2,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=ProsperRating..Alpha.)) +
  scale_color_brewer(type = 'seq',palette='Reds',limits=c("AA","A","B","C","D",
                                                          "E","HR")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Term) +
  xlab('Credit Score')

```


### Description Two
> Shorter term listings have lower BorrowerRate. The rates are further layered \
by ProsperRating..Alpha. We can also see moderate negative correlation between\
BorrowerRate and CreditScore.



### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Summary3}
ggplot(aes(x=StatedMonthlyIncome,y=BorrowerRate),
       data=subset(loan_data,as.character(ProsperScore)!="")) +
  geom_jitter(aes(color=ProsperRating..Alpha.)) +
  scale_color_brewer(type = 'seq',palette='Blues',limits=c("AA","A","B","C","D",
                                                           "E","HR")) +
  xlim(0,30000) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Term) +
  xlab('Stated Monthly Income ($)')

```

### Description Three

> In comparison with Plot2, the positive correlation between borrower rate and\
stated monthly income is weaker than that between borrower rate and borrower's\
credit score although one would expect that borrower's income should play a\
critical role in determining the riskness of a loan.

------

# Reflection

The loan data set has 113,937 records across 81 variables. There is a clear gap
during pre-2009 and post-2009, possibly due to the Financial Crisis in late
2008. The focus of this EDA is on the loan listed post-2009.

My objective is to determine what determines the interest rate a borrower has
to pay for his loan. I suspect that the factors would include things like his
credit score, his past settlement history, amount of debts he hold, his income,
his employment, and loan term.

Firstly, I examined many variables, plotting out their histograms and see their
staitstics. Next I explored the relationships between two variables. As
expected, there is a strong positive correlation between interest rate and the
credit rating of a loan, but my main question is what determines which credit
rating a loan will receive. The scatterplot between interest rate and
borrower's credit score shows a moderate positive correlation. However, other
variables such as number of open credit lines, number of open revolving
accounts, stated monthly income, and loan size do not show clear correlation.

When I used facet_wrap to see whether the loan term affects interest rates, I
found that the 12-month loan has lower interest rate as expected. However, I am
surprised to find that there is almost no difference in the median interest
rates of 36-month and 60-month loans.

When I tried to find a linear model to predict a borrower's interest rate, only
credit rating of the loan and borrower's credit score significantly contribute
to the R-squared value. Other factors' contributions are negligible.

For future exploration, I may have to include things like loan categories,
occupations, and collaterals in order to better determine the required interest
rates.
